---
title: "Phylogenetic Tree Inspection & Rooting"
author: "Andrea Martinez"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document: 
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      # Always relevant to the document directory 
                      # send any figure output to this folder 
                      fig.path = "../figures/03_Phylogenetic_Tree/",
                      warning = FALSE) 
```
# Goals 

1. Load the fastree unrooted tree.  
2. Add tree to phyloseq object.  
3. Visualize and inspect tree with ggtree. 
4. Prune ASVs, if needed.  
5. Root our tree. 
6. Combine new tree with a phyloseq object. 
7. Save 2 phyloseq objects: 1. Unrooted tree phyloseq object, 2. Rooted tree phyloseq object. 

# Before you start

## Set my seed 
```{r set-seed}
# Any number can be chosen 
set.seed(238428)
```

## Load Packages 
```{r load-packages}
pacman::p_load(tidyverse, phyloseq, ggtree, phytools,
               install = FALSE)
```
## Load Data files 
```{r load-data}
# Preprocessed phyloseq object 
load("data/02_PreProcessing/raw_preprocessed_physeq.RData")
raw_preprocessed_physeq

# Load in the tree! 
unrooted_tree <- read.tree("data/03_Phylogenetic_Tree/ASVs_unrooted.tree")
unrooted_tree
str(unrooted_tree)
```
# Merge Phyloseq 
```{r merge-physeq}
# Intuition check 
stopifnot(ntaxa(raw_preprocessed_physeq) == ntaxa(unrooted_tree))
# Merge the tree with the phyloseq object 
unrooted_physeq <- 
  merge_phyloseq(raw_preprocessed_physeq, unrooted_tree)
unrooted_physeq
```

# Plot Tree with `ggtree`

```{r plot-tree-unrooted}
# Make a basic tree
kingdom_tree <- 
  ggtree(unrooted_physeq) + 
  # color tips by kingdom 
  geom_tippoint(mapping = aes(color = Kingdom)) + 
  scale_color_manual(values = c("goldenrod1", "cornflowerblue", "grey")) +
  # Add title 
  labs(title = "Unrooted Tree") + 
  #move the legend to the bottom 
  theme(legend.position = "bottom"); kingdom_tree

kingdom_node_tree <- 
  kingdom_tree + 
  # Add the node label 
  geom_text(aes(label=node), hjust= -0.5, vjust = -0.3, size = 2)
kingdom_node_tree
```

# Evaluate Long Branches 

```{r eval-long-branch}
# View a specific clade 
# Zoom in on origin tree: Node 2537
viewClade(kingdom_node_tree + 
          labs(title = "Unrooted Tree: Node 59348"), 
          node = 59348)
# Looks okay to me.

viewClade(kingdom_node_tree + 
          labs(title = "Unrooted Tree: Node 11351") + 
          geom_text(aes(label=ASV)), 
          node = 11351)
# ASV_18392
```

```{r check-taxonomy}
# Let's make a note of the the taxonomy and the ASV Sequence.... 
# Let's make a note of the the taxonomy and the ASV Sequence.... 
unrooted_physeq %>%
  subset_taxa(., ASV == "ASV_18392") %>%
  tax_table() %>%
  data.frame()

# Let's also check the counts of the ASV 
unrooted_physeq %>%
  subset_taxa(., ASV == "ASV_18392") %>%
  otu_table() %>%
  data.frame() %>%
  colSums()

# Present in 1 out of 35 samples with a read count of 2.

# Search sequence on NCBI BLAST.
```

ASV_18392 shares 89.17% identity with Clostridium bovifaecis strain BXX. Sequences with higer similarites were labeled only as "Uncultured bacterium clone" Clostridium bovifaecis is an acetogenic bacterium found in cow manure, which could reasonably be present in an agricultural setting. However, the <90% shared identity is not high enough to confidently identify the bacteria to the species or even to the genus level. Despite its unknown identity, due to there being very counts of this ASV and its presence in only a single sample, it can be pruned from the tree.

# Prune ASV_18392
```{r prune-ASV18392}
# Function from Joey McMurdie: https://github.com/joey711/phyloseq/issues/652
pop_taxa = function(physeq, badTaxa){
  allTaxa <-  taxa_names(physeq)
  allTaxa <- allTaxa[!(allTaxa %in% badTaxa)]
  return(prune_taxa(allTaxa, physeq))}

# Let's use the pop_taxa function :) 
# Recreate a phyloseq object without ASV_18392
unrooted_physeq_rm18392 <- 
  unrooted_physeq %>%
  pop_taxa(., "ASV_18392")

# Intuition Check
ntaxa(unrooted_physeq) - ntaxa(unrooted_physeq_rm18392) 

# Visually inspect 
ggtree(unrooted_physeq_rm18392) + 
  geom_text(aes(label=node)) +
  geom_tippoint(mapping = aes(color = Kingdom))
```

# Midroot Tree
```{r midroot-tree}
# Is the tree rooted?
new_unrooted_tree <-phy_tree(unrooted_physeq_rm18392)
is.rooted(new_unrooted_tree)

# Let's midpoint root the tree
midpoint_rooted_tree <- midpoint.root(new_unrooted_tree)

# Is the new tree rooted?
is.rooted(midpoint_rooted_tree)

# Assign to a new phyloseq object: merging subsetted phyloseq with the new rooted tree

# 1. Create the phyloseq object without ASV_18392
physeq_rm18392 <- 
  raw_preprocessed_physeq %>%
  subset_taxa(ASV != "ASV_18392")

# Merge tree with the new physeq_rm18392
midroot_physeq_rm18392 <- 
  merge_phyloseq(physeq_rm18392, midpoint_rooted_tree)
midroot_physeq_rm18392

# Quick inspection of tree 
ggtree(midroot_physeq_rm18392) + 
  geom_tippoint(mapping = aes(color = Kingdom))
```

# Save to a new phyloseq object
```{r save-physeq}
# Save both phyloseq objects with our tree object to one .RData file 
save(list = c("unrooted_physeq_rm18392", "midroot_physeq_rm18392"),
     file = "data/03_Phylogenetic_Tree/phytree_preprocessed_physeq.RData")
```

# Session Information 
```{r session-info}
# Ensure reproducibility 
devtools::session_info()
```